---
title: "EDS241: Assignment 3"
author: "Charles Hendrickson"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)


# load packages
packages=c("stargazer", "here", "tidyr", "dplyr","stringr", "janitor", 
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble", "openxlsx", "estimatr", "car", "readxl")

# Load estimatr package
library(estimatr)

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

#devtools::install_github('rstudio/rmarkdown')
options(scipen=10) # not scientific notation

```

Load the SMOKING_EDS241.csv data 
```{r}
# load data 
smoking_data <- read.csv("SMOKING_EDS241.csv")

```

\noindent (a) What is the unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers? Under what assumption does this correspond to the average treatment effect of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical evidence for or against this hypothesis.

For the last part of (a), I would regress your favorite covariate on the smoking status of mothers. For example, think of regressing meduc ~ tobacco. Is the mean difference in the education level of smoking and non-smoking mothers statistically different from zero? What does that say about the required assumption to interpret the unadjusted mean difference as causal? Let me know if this tip is still too cryptic.

**The unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers is -244.5 grams.**

**Under the "treatment ignorability" assumption, this corresponds to the average treatment effect of maternal smoking during pregnancy on infant birth weight**

**When regressing the mother's education level (meduc) on the indicator for maternal smoking (tobacco), the mean difference in the education level of smoking and non-smoking mothers is -1.318 grams which is statistically significant becuase the p-value < 2.2e-16, which is much lower than the 5% significance level. Therefore, we can interpret the unadjusted mean difference as causal.**


```{r}
# Regress birth weight of infant in grams (birthwgt) on the indicator for maternal smoking (tobacco)
summary(lm_robust(birthwgt ~ tobacco, data = smoking_data))
```

```{r}
# Regress the mother's education level (meduc) on the indicator for maternal smoking (tobacco)
summary(lm_robust(meduc ~ tobacco, data = smoking_data))
```

\noindent (b) Assume that maternal smoking is randomly assigned conditional on the observable
covariates listed above. Estimate the effect of maternal smoking on birth weight using a linear
regression. Report the estimated coefficient on tobacco and its standard error.

On (b), simply include all of the covariates linearly. We are making a functional form assumption on the linear impact of covariates on the outcome.

**The estimated coefficient on tobacco is -228.073 grams and the standard error is 4.2768.**

```{r}
summary(lm_robust(birthwgt ~ tobacco + anemia + diabete + alcohol + mblack + first + mage + meduc, data = smoking_data))
```


\noindent (c) Use the exact matching estimator to estimate the effect of maternal smoking on birth weight. For simplicity, consider the following covariates in your matching estimator: create a 0-1
indicator for mother's age (=1 if mage>=34), and a 0-1 indicator for mother's education (1 if
meduc>=16), mother's race (mblack), and alcohol consumption indicator (alcohol). These 4
covariates will create 2*2*2*2 = 16 cells. Report the estimated average treatment effect of
smoking on birthweight using the exact matching estimator and its linear regression analogue
(Lecture 6, slides 12-14).

For (c), once you have your 4 dummy variables, you can create a group variable g using the paste0() function. For example, mutate(g = paste0(d1,d2,d3,d4)) . The resulting g will include all potential observed combinations of the 4 dummy variables in the data. You can then control for factor(g) in the regression model. This is much simpler than interacting all the dummy variables as we talked about in OHs today.
For (c), to calculate the exact matching estimator, use this g grouping variable and the code from lines 76 to 97 in the TIA.R script on gauchospace. In this case, Y = birthwgt, X = g, D = tobacco. Since we observe Y1 or Y0, you can ignore line 77.

```{r}

```




































